Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-22 manuscript/main.qmd
remove degC behind brackets

The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig. 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"`--`r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"`--`r tdiff_lat90$"97.5%"`)&deg;C higher. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"`--`r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"`--`r eocene_gradient_polar_circle$"97.5%"`)&deg;C.

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient--`r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"`--`r eeco_sdy$"97.5%"`)&deg;C--which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`. This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"`--`r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern. A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"`--`r eoceneproxy_global_diff$"97.5%"`)&deg;C. The median modelled temperature is higher near the equator and in high latitudes when excluding the ecological proxies, with a flattened median gradient of `r eocene_proxy_gradient$"50%"`&deg;C (Fig. S2). In contrast, including ecological proxies, but widening the uncertainty around the low-latitude ecological proxy data does not significantly change the resulting gradient (Fig. S3).

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"`--`r eeco_south_north_global_diff$"97.5%"`)&deg;C. The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"`--`r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig. S4).

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7--21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but a lot less steep when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg. Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world. Likewise, climate simulations using general circulation or Earth system models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]

>>>>

The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig. 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"`&deg;C (95% CI: `r tdiff_lat0$"2.5%"`--`r tdiff_lat0$"97.5%"`&deg;C) higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"`&deg;C (`r tdiff_lat90$"2.5%"`--`r tdiff_lat90$"97.5%"`&deg;C) higher. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`&deg;C (`r eocene_gradient$"2.5%"`--`r eocene_gradient$"97.5%"`&deg;C) for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"`&deg;C (`r eocene_gradient_polar_circle$"2.5%"`--`r eocene_gradient_polar_circle$"97.5%"`&deg;C).

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient--`r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"`--`r eeco_sdy$"97.5%"`)--which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`. This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

The early Eocene GMST is estimated at `r eotemp_global$"50%"`&deg;C (`r eotemp_global$"2.5%"`--`r eotemp_global$"97.5%"`&deg;C), `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern. A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"`&deg;C (`r eoceneproxy_global_diff$"2.5%"`--`r eoceneproxy_global_diff$"97.5%"`&deg;C). The median modelled temperature is higher near the equator and in high latitudes when excluding the ecological proxies, with a flattened median gradient of `r eocene_proxy_gradient$"50%"`&deg;C (Fig. S2). In contrast, including ecological proxies, but widening the uncertainty around the low-latitude ecological proxy data does not significantly change the resulting gradient (Fig. S3).

Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres. Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"`&deg;C (`r eeco_south_north_global_diff$"2.5%"`--`r eeco_south_north_global_diff$"97.5%"`&deg;C). The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI of that difference spans `r eocene_north_south_gradient_diff$"2.5%"`--`r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig. S4).

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7--21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but a lot less steep when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`&deg;C, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg. Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world. Likewise, climate simulations using general circulation or Earth system models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]
